Natural selection in the Health and Retirement Study

### 2023-09-16 hrs-selection.qmd
Create a single dataframe for both ethnicities, split when necessary

```{r setup}

@@

calc_income <- function (x) {
  ages <- x |> select(matches("r\\d+agey_e"))
  income <- x |> select(matches("r\\d+iearn"))
  income <- rowMeans(income, na.rm = TRUE)
  income_r <- resid(lm(income ~ factor(x$rabyear)))
  income_r <- income_r - min(income_r) + 1

@@

@@

  rand <- inner_join(rand_orig, pgen,
                     join_by(hhid, pn),
                     unmatched = "drop",
                     relationship = "one-to-one")

  rand <- rand |> left_join(rand_fam, 
                            by = join_by(hhid, pn),
                            unmatched = "drop",
                            relationship = "one-to-one")
  rand <- rand |>
    mutate(.by = rabyear,
      mean_raevbrn = mean(raevbrn, na.rm = TRUE),
      rlrs = raevbrn/mean_raevbrn
    ) |>
    mutate(
      income_resid = calc_income(rand),
      income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
      income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
      # as.numeric avoids haven.labelled issues:
      raedyrs = as.numeric(raedyrs),
      edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
      # wave 10 is 2010
      # table(rands$white$r10mstat)
       # 1                 1.married
       # 2   2.married,spouse absent
       # 3               3.partnered
       # 4               4.separated
       # 5                5.divorced
       # 6      6.separated/divorced
       # 7                 7.widowed
       # 8           8.never married
      married = factor(r10mstat == 1, 
                       levels = c(FALSE, TRUE),
                       labels = c("Other", "Married")),
    r10sibs = r10livsib + r10sbdied,
    agefb = r10agey_e - h10ageokid,
    agefb = ifelse(agefb < 12, NA_real_, agefb),
    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
  )

  # did this to check if white larger sample size drove result sizes; 
  # it didn't:
  # rand <- rand |> slice_sample(n = 3100)

@@

  rands[[eth]] <- rand
}

@@

```{r stat-regressions}

@@

cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
for (eth in eths) {
  rand <- rands[[eth]]

  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
  mod_pgs[[eth]] <- map(form_pgs, 
                        \(x) lm(x, data = rand, weights = weight)
                        )

@@

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                              \(x) lm(x, data = rand, weights = weight)
                              )

@@

  form_pgs_sibs <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "r10sibs"))
  mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
                             \(x) lm(x, data = rand, 
                                     weights = weight/(1 + r10sibs)))

@@

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
        \(x) lm(x, data = rand, weights = weight)
        )

@@

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
        \(x) lm(x, data = rand, weights = weight)
        )

@@

  form_pgs_x_mar <- paste0(pgs, ":married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
        \(x) lm(x, data = rands[[eth]], weights = weight)
        )

@@

  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "raevbrn"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
        \(x) lm(x, data = rands[[eth]], weights = weight)
        )

@@

}

@@

cor_fert_edu_phen <- list()
for (eth in eths) {
  cor_fert_edu_phen[[eth]] <- lm(raevbrn ~ raedyrs, data = rands[[eth]],
                                 weights = weight) |> 
    coef() |> 
    pluck("raedyrs")
}

@@

n_reps <- 599
cors_boot <- list()
for (eth in eths) {
  rand_short <- rands[[eth]] |> 
    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight, live_sibs)
  cors_boot[[eth]] <-
    replicate(n_reps, {
      rand <- rand_short |> 
              slice_sample(prop = 1, replace = TRUE)

      coef_pgs_fert <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
        map(\(x) lm(x, data = rand, weights = weight)) |> 
        map_dbl(\(x) coef(x)[[2]])

      coef_pgs_sibs <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "live_sibs")) |> 
        map(\(x) lm(x, data = rand, weights = weight/(live_sibs + 1))) |> 
        map_dbl(\(x) coef(x)[[2]])

      coef_pgs_edyrs <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
        map(\(x) lm(x, data = rand, weights = weight)) |> 
        map_dbl(\(x) coef(x)[[2]])

@@

for (eth in eths) {
  rand <- rands[[eth]]
  rand_short <- rands[[eth]] |> 
    select(
      all_of(pgs), all_of(pcs), 
      raevbrn, raedyrs, weight, income_med, edu, married
    )

@@

```{r}
#| label: fig-cohorts
#| fig-cap: Distribution of birth years for the sample
#| fig-subcap:
#|   - ""
#|   - ""
#| layout-nrow: 2
#| column: page

for (eth in eths) {
  ggp <- ggplot(rands[[eth]], aes(rabyear)) + 
    geom_bar() +
    theme_minimal() +
    labs(x = "Birth year", y = "Count", title = titles[[eth]])
  print(ggp)
}
```

@@

>>>>

```{r}
#| label: setup

@@

calc_income <- function (x, eth) {
  x <- x |> filter(ethnicity == eth)
  ages <- x |> select(matches("r\\d+agey_e"))
  income <- x |> select(matches("r\\d+iearn"))
  income <- rowMeans(income, na.rm = TRUE)
  income_r <- resid(lm(income ~ factor(x$rabyear)))
  income_r <- income_r - min(income_r) + 1

@@

calc_n <- function (mods, eths = c("black", "white"), format = "(N = %s)") {
  nobs <- map_dbl(eths, \(x) nobs(mods[[x]][[1]]) ) 
  if (! is.null(format)) nobs <- sprintf(format, nobs)
  names(nobs) <- eths
  nobs
}

@@

  rands[[eth]] <- inner_join(rand_orig, pgen,
                     join_by(hhid, pn),
                     unmatched = "drop",
                     relationship = "one-to-one")
}

rand <- list_rbind(rands, names_to = "ethnicity")

rand <- rand |> left_join(rand_fam, 
                          by = join_by(hhid, pn),
                          unmatched = "drop",
                          relationship = "one-to-one")
rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + r10sbdied,
    agefb = r10agey_e - h10ageokid,
    agefb = ifelse(agefb < 12, NA_real_, agefb),
    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
  )

@@

@@

```{r}
#| label: stat-regressions
#| cache: true

@@

cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
cor_fert_edu_phen <- list()

for (eth in eths) {
  r_eth <- rand |> filter(ethnicity == eth)

  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
  mod_pgs[[eth]] <- map(form_pgs, 
                        \(x) lm(x, data = r_eth, weights = weight)
                        )

@@

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                              \(x) lm(x, data = r_eth, weights = weight)
                              )

@@

  form_pgs_sibs <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "r10sibs"))
  mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
                             \(x) lm(x, data = r_eth, 
                                     weights = weight/(1 + r10sibs)))

@@

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
        \(x) lm(x, data = r_eth, weights = weight)
        )

@@

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
        \(x) lm(x, data = r_eth, weights = weight)
        )

@@

  form_pgs_x_mar <- paste0(pgs, ":married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
        \(x) lm(x, data = r_eth, weights = weight)
        )

@@

  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "raevbrn"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
        \(x) lm(x, data = r_eth, weights = weight)
        )

@@

  cor_fert_edu_phen[[eth]] <- 
    lm(raevbrn ~ raedyrs, data = r_eth, weights = weight) |> 
    coef() |> 
    pluck("raedyrs")
}

@@

@@

n_reps <- 599
cors_boot <- list()
for (eth in eths) {
  rand_short <- rand |> 
    filter(ethnicity == eth) |> 
    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight, r10sibs)
  cors_boot[[eth]] <-
    replicate(n_reps, {
      r_boot <- rand_short |> 
              slice_sample(prop = 1, replace = TRUE)

      coef_pgs_fert <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
        map(\(x) lm(x, data = r_boot, weights = weight)) |> 
        map_dbl(\(x) coef(x)[[2]])

      coef_pgs_sibs <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "r10sibs")) |> 
        map(\(x) lm(x, data = r_boot, weights = weight/(r10sibs + 1))) |> 
        map_dbl(\(x) coef(x)[[2]])

      coef_pgs_edyrs <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
        map(\(x) lm(x, data = r_boot, weights = weight)) |> 
        map_dbl(\(x) coef(x)[[2]])

@@

for (eth in eths) {
  rand_short <- rand |> 
    filter(ethnicity == eth) |> 
    select(
      all_of(pgs), all_of(pcs), 
      raevbrn, raedyrs, weight, income_med, edu, married
    )

@@

```{r}
#| label: fig-cohorts
#| fig-cap: Distribution of birth years for the sample

rand |> 
    mutate(
      Ethnicity = fct_relevel(titles[ethnicity], "White", "Black")
    ) |> 
    ggplot(aes(rabyear, fill = Ethnicity)) + 
      geom_bar(position = "identity") +
      labs(x = "Birth year", y = "Count") +
      scale_fill_manual(values = c(Black = "orange2", White = "steelblue4")) +
      theme_minimal() +
      theme(legend.title = element_blank())

@@

for (eth in eths) {}
